Tulai

### 2023-09-29 paper/paper.qmd
avoid function name conflicts with ::, replace PCA ellipses with convex hulls

```{r}
# Read the data
tl <- read_csv(here("data/raw-data/Tulai_Lithic_Assamblage.csv"))

# this file is prepared from the google sheet here
# https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0
# and simplified using code in the scraps.R file
excavation_data <- read_csv(here("data//raw-data/excavation_data.csv"))
```

@@

tl4 <- tl4 %>%
  # BM: I updated this to use across
  mutate(across(everything(), 
                ~ifelse(is.na(.x)|.x == ""|.x == "-", NA, .x))) %>% 
  #SS:  remove any asterisks, whitespace, or *, ?, or !. is that wrong?
   mutate(across(everything(),
                 ~str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", ""))) %>%
  ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", 
                   "Parallel", 
                   "Sub-Parallel", 
                   "Parallel-Subprarallel",
                   "Scaled", 
                   "Sub-Paralel", 
                   "Sub-Parallell", 
                   "Sub-Paallel",
                   "Sub-Paralell", 
                   "Sub-Prallel", 
                   "L:semiparallel/R:scaled",
                   "L:semiparallel/R: Scaled",
                   "Semiparallel", 
                   "Semi-Parallel", 
                   "Scaled/Sub-Parallel",
                   "Seb-Parallel", 
                   "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
    Retouch =     if_else(Retouch == "0", NA_character_, Retouch),
    SHINE =       if_else(SHINE == "0", NA_character_, SHINE),
    shine2 =      if_else(shine2 == "0", NA_character_, shine2),
    Blank2 =      if_else(Blank2 == "0", NA_character_, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

@@

tl_final <- tl4 %>%
  rename(
    blank =                Blank2,
   `sickle shine` =       shine2,
    length =              `Length (mm)` ,
    width =               `Width (mm)`,
    thickness =           `Thickness (mm)`, 
    typology =             Typology,
    utilization =          Utilization,
    retouch =              Retouch,
    blank =                Blank2, 
   `blank part` =         `Blank_part`,
   `bulb thickness` =     `Bulb Thickness (mm)`, 
   `platform thickness` = `Platform Thickness (mm)`, 
   `platform length` =    `platform Length (mm)`, 
    eraillure =            Eraillure , 
    cortex =               `Cortex (%)`)  %>%
  mutate(area_unit = paste0(toupper(area), "_", chronological_unit)) %>%
  relocate(area_unit, .after = chronological_unit)   %>%
  relocate(grouped_level, .after = area_unit)   %>%
  select(lithic_id, 
         TA, 
         area, 
         chronological_unit,
         area_unit,
         grouped_level,
         cortex,
         depth, upper, lower, midpoint, 
        `length`, `width`, `thickness`, 
         typology, 
         utilization, 
        `sickle shine`,
         retouch, 
        `typology-tool-core`,
        `tool-typology`,
         blank, `blank part`, 
        `core-typology`, `core-fragment`,`core-technology`, 
        cortex)  %>% 
# Update the 'typology-tool-core' column based on 'Retouch'
  mutate(
    `typology-tool-core` = case_when(
      retouch == 1 ~ "tool",           
      utilization == 1 ~ "tool",
      `sickle shine` == 1 ~ "tool",
      TRUE ~ `typology-tool-core`      
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
      utilization == 1 ~ "utilized tool",
      `sickle shine` == 1 ~ "sickle shine",
      TRUE ~ `tool-typology`               
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
     # If Retouch is 1 and tool-typology is NA
       retouch == 1 & is.na(`tool-typology`) ~ "retouched piece",
       `sickle shine` == 1 & is.na(`tool-typology`) ~ "sickle shine",

@@

```{r}
# Create a summary table
summary_table_area <- tl_final %>%
  group_by(area) %>%
  summarise(
   `Cores (n)` = sum((!is.na(`core-typology`) & `core-typology` != "") |
                      (!is.na(core) & core != ""), na.rm = TRUE),  # Summing variables from both columns
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag) &
                                            corefrag != ""),
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

@@

summary_table_area_filtered <- 
  summary_table_area %>%
  filter(area %in% c("ta", "d1", "tp1"))

@@

write_csv(summary_table_area_filtered,
          here("data/derived-data/tbl-001-summary_table_area_filtered.csv"))

@@

total_row <- summary_table_tool_area %>%
  select(-tooltypegroup) %>%
  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
  mutate(tooltypegroup = "Total")

@@

write_csv(summary_table_tool_area,
          here("data/derived-data/tbl-002-summary_table_tool_area.csv"))

@@

# write the percentage table to a CSV file
write_csv(summary_table_tool_area_percentage,
          here("data/derived-data/tbl-002-summary_table_tool_area_percentage.csv"))

@@

tl_final_excavation_data_volume_artefact_counts <- 
tl_final_excavation_data_volume %>% 
  group_by(label2) %>% 
   summarise(
    blade_count =              sum(blank == "blade", na.rm = TRUE),
    flake_count =              sum(blank == "flake", na.rm = TRUE),
    bladelet_count =           sum(blank == "bladelet", na.rm = TRUE),
    tool_count =               sum(retouch == 1, 
                                   na.rm = TRUE) + sum(utilization == 1, 
                                                       na.rm = TRUE),
    core_count =                sum(!is.na(`core-typology`), na.rm = TRUE) + sum(!is.na(core), na.rm = TRUE),

@@

ggsave(here("figures/fig-wabi-plot.jpg"),
       h = 6, 
       w = 8)

@@

ggsave(here("figures/fig-contribution-plots.jpg"),
       h = 6, 
       w = 8)

@@

write_csv(eigen_df,
          here("data/derived-data/tbl-003-pca-eigenvalues.csv"))

@@

# get excavation area so we can use it for grouping
excavation_area <- 
  str_extract(sum_depth_sheet_subset$label2, 
              "[^_]+")

# Generate the base biplot
biplot_base <- fviz_pca_biplot(res.pca,
                               axes = c(1, 2),
                               labelsize = 2,
                               label = "var",
                               addEllipses = TRUE,
                               col.ind = excavation_area,
                               title = ""
                              )

# Modify row names of coordinates
rownames(res.pca$ind$coord) <- sum_depth_sheet_subset$label2

pca_label_text <- 
res.pca$ind$coord %>% 
  as_tibble() %>% 
  mutate(labels = sum_depth_sheet_subset$label2,
         area =   excavation_area)

# Add geom_text_repel() to the base plot
biplot_modified <- biplot_base +
                   geom_text_repel(data = pca_label_text,
                                   aes(x = Dim.1, 
                                       y = Dim.2, 
                                       label = labels,
                                       colour = area),
                                   segment.color = "grey80",
                                   size = 2,
                                   force = 10
                                  ) +
  labs(colour = "",
       shape = "",
       fill = "") +
  theme(legend.position = c(0.9, 0.2))

@@

ggsave(here("figures/fig-pca-biplot.jpg"),
       h = 6, 
       w = 8)

>>>>

```{r}
# Read the data
tl <- read_csv(here::here("data/raw-data/Tulai_Lithic_Assamblage.csv"))

# this file is prepared from the google sheet here
# https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0
# and simplified using code in the scraps.R file
excavation_data <- read_csv(here::here("data//raw-data/excavation_data.csv"))
```

@@

tl4 <- tl4 %>%
  # BM: I updated this to use across
  mutate(across(everything(), 
                ~ifelse(is.na(.x)|.x == ""|.x == "-", NA, .x))) %>% 
  #SS:  remove any asterisks, whitespace, or *, ?, or !. is that wrong?
   mutate(across(everything(),
                 ~str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", ""))) %>%
  ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", 
                   "Parallel", 
                   "Sub-Parallel", 
                   "Parallel-Subprarallel",
                   "Scaled", 
                   "Sub-Paralel", 
                   "Sub-Parallell", 
                   "Sub-Paallel",
                   "Sub-Paralell", 
                   "Sub-Prallel", 
                   "L:semiparallel/R:scaled",
                   "L:semiparallel/R: Scaled",
                   "Semiparallel", 
                   "Semi-Parallel", 
                   "Scaled/Sub-Parallel",
                   "Seb-Parallel", 
                   "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA, Utilization),
    Retouch =     if_else(Retouch ==     "0", NA, Retouch),
    SHINE =       if_else(SHINE ==       "0", NA, SHINE),
    shine2 =      if_else(shine2 ==      "0", NA, shine2),
    Blank2 =      if_else(Blank2 ==      "0", NA, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

@@

tl_final <- tl4 %>%
  dplyr::rename(
    blank =                Blank2,
   `sickle shine` =        shine2,
    length =              `Length (mm)` ,
    width =               `Width (mm)`,
    thickness =           `Thickness (mm)`, 
    typology =             Typology,
    utilization =          Utilization,
    retouch =              Retouch,
    blank =                Blank2, 
   `blank part` =         `Blank_part`,
   `bulb thickness` =     `Bulb Thickness (mm)`, 
   `platform thickness` = `Platform Thickness (mm)`, 
   `platform length` =    `platform Length (mm)`, 
    eraillure =            Eraillure , 
    cortex =               `Cortex (%)`)  %>%
  mutate(area_unit = paste0(toupper(area), "_", chronological_unit)) %>%
  relocate(area_unit, .after = chronological_unit)   %>%
  relocate(grouped_level, .after = area_unit)   %>%
  select(lithic_id, 
         TA, 
         area, 
         chronological_unit,
         area_unit,
         grouped_level,
         cortex,
         depth, upper, lower, midpoint, 
        `length`, `width`, `thickness`, 
         typology, 
         utilization, 
        `sickle shine`,
         retouch, 
        `typology-tool-core`,
        `tool-typology`,
         blank, `blank part`, 
        `core-typology`, `core-fragment`,`core-technology`, 
        cortex)  %>% 
# Update the 'typology-tool-core' column based on 'Retouch'
  mutate(
    `typology-tool-core` = case_when(
      retouch == 1 ~ "tool",           
      utilization == 1 ~ "tool",
      `sickle shine` == 1 ~ "tool",
      TRUE ~ `typology-tool-core`      
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
      utilization == 1 ~ "utilized tool",
      `sickle shine` == 1 ~ "sickle shine",
      TRUE ~ `tool-typology`               
    )
  ) %>% 
  mutate(
    `tool-typology` = case_when(
     # If Retouch is 1 and tool-typology is NA
       retouch == 1 & is.na(`tool-typology`) ~ "retouched piece",
       `sickle shine` == 1 & is.na(`tool-typology`) ~ "sickle shine",

@@

```{r}
# Create a summary table
summary_table_area <- tl_final %>%
  group_by(area) %>%
  dplyr::summarise(
   `Cores (n)` = sum((!is.na(`core-typology`) & `core-typology` != "") |
                      (!is.na(core) & core != ""), na.rm = TRUE),  # Summing variables from both columns
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag) &
                                            corefrag != ""),
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

@@

summary_table_area_filtered <- 
  summary_table_area %>%
  dplyr::filter(area %in% c("ta", "d1", "tp1"))

@@

write_csv(summary_table_area_filtered,
          here::here("data/derived-data/tbl-001-summary_table_area_filtered.csv"))

@@

total_row <- summary_table_tool_area %>%
  select(-tooltypegroup) %>%
  dplyr::summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
  mutate(tooltypegroup = "Total")

@@

write_csv(summary_table_tool_area,
          here::here("data/derived-data/tbl-002-summary_table_tool_area.csv"))

@@

# write the percentage table to a CSV file
write_csv(summary_table_tool_area_percentage,
          here::here("data/derived-data/tbl-002-summary_table_tool_area_percentage.csv"))

@@

tl_final_excavation_data_volume_artefact_counts <- 
tl_final_excavation_data_volume %>% 
  group_by(label2) %>% 
   dplyr::summarise(
    blade_count =              sum(blank == "blade", na.rm = TRUE),
    flake_count =              sum(blank == "flake", na.rm = TRUE),
    bladelet_count =           sum(blank == "bladelet", na.rm = TRUE),
    tool_count =               sum(retouch == 1, 
                                   na.rm = TRUE) + sum(utilization == 1, 
                                                       na.rm = TRUE),
    core_count =                sum(!is.na(`core-typology`), na.rm = TRUE) + sum(!is.na(core), na.rm = TRUE),

@@

ggsave(here::here("figures/fig-wabi-plot.jpg"),
       h = 6, 
       w = 8)

@@

ggsave(here::here("figures/fig-contribution-plots.jpg"),
       h = 6, 
       w = 8)

@@

write_csv(eigen_df,
          here::here("data/derived-data/tbl-003-pca-eigenvalues.csv"))

@@

# get excavation area so we can use it for grouping
excavation_area <- 
  str_extract(sum_depth_sheet_subset$label2, 
              "[^_]+")
pca_data.pca <- prcomp(pca_data, scale. = TRUE)

biplot_modified <- 
ggbiplot(
  pca_data.pca,
  obs.scale = 1,
  var.scale = 1,
  groups = excavation_area,
  ellipse = FALSE,
  circle = TRUE) +
  stat_chull(aes(color = excavation_area,
                 fill = excavation_area),
             alpha = 0.1,
             geom = "polygon") +
  scale_colour_brewer(palette = "Set1",
                      name = '',
                      guide = 'none') +
  scale_fill_brewer(palette = "Set1",
                    name = '')  +
  theme_minimal()

@@

ggsave(here::here("figures/fig-pca-biplot.jpg"),
       h = 6, 
       w = 8)
